\section{Modified Beam Search with Faithfulness-Aware Reranking}
\label{ch:beam_search}

\subsection{Methodology}

We integrate the two faithfulness signals validated in
Sections~\ref{sec:factscore_turbo} and~\ref{sec:lookback_lens} into the
decoding step of the target generator.  At each step the decoder expands the
top-$k$ next-token candidates per beam and reranks the resulting
$(k \cdot W)$ candidates by

\begin{equation}
\mathrm{score}(b) =
\frac{\sum_{t=1}^{L_b} \log p_t}{L_b^{\alpha}}
\;+\; \lambda_{\text{LL}}\, \mathrm{LL}(b)
\;+\; \mathbf{1}[\text{sentence end}] \cdot \lambda_{\text{FS}}\, \mathrm{FS}(b),
\label{eq:guided_score}
\end{equation}

where $L_b$ is the response length of beam $b$, $\alpha = 0.7$ is the GNMT
length-normalisation exponent~\cite{wu2016google},
$\mathrm{LL}(b) = 1 - p_{\text{halluc}}(\text{features}(b))$ is the
groundedness signal from the trained Lookback Lens classifier, and
$\mathrm{FS}(b)$ is the FActScore-Turbo verdict on the most recent sentence
finished by beam $b$.  When no atomic facts can be extracted at a sentence
boundary, $\mathrm{FS}(b)$ defaults to a neutral $0.5$ rather than the empty-
sentence trivial $1.0$, neutralising the degenerate \emph{say nothing}
strategy that arises when a bonus is summed into beam-search scores.

The lookback ratio features are obtained \emph{from the same forward pass}
that produced the next-token logits, so the only additional cost per step is
a single per-beam logistic-regression evaluation.  The FActScore signal is
sentence-level: it triggers only on tokens containing $\{., !, ?\}$, keeping
its $\sim$6\,s/call cost under control even with a beam width of four.

\subsection{Implementation}

The custom decoding loop is implemented in \texttt{src/guided\_beam\_search.py}.
A reusable helper \texttt{LookbackRatioExtractor.compute\_from\_attentions}
exposes the pure attention\,$\to$\,feature reduction so the loop can avoid a
second forward pass.  Two judges from different model families
(\texttt{qwen2.5:14b} and \texttt{llama3.1:8b}) score the final outputs on
three independent dimensions.  All code paths are covered by 45 unit tests
that pass without external services.

\subsection{Pilot Experimental Setup}

The headline experiment is intended for \texttt{Qwen/Qwen2.5-7B-Instruct},
but the necessary model weights had not finished downloading at the time of
this report.  We therefore present a \emph{pilot run} on
\texttt{facebook/opt-125m}, the same model used for the Lookback Lens
classifier baseline (Section~\ref{sec:lookback_lens}, AUC$=0.625$).  The
pilot validates the implementation end-to-end on real RAGTruth contexts and
documents the methodology that the Qwen-7B run will follow once the
download completes.

\begin{itemize}
\item \textbf{Generator}: \texttt{facebook/opt-125m} (12 layers, 12 heads,
$\sim$125\,M params), float32 on MPS, beam width $W=2$, $\alpha=0.7$,
$\max_{\text{new tokens}}=48$.
\item \textbf{Lookback Lens classifier}: re-used from the baseline
experiment (\texttt{experiments/lookback\_lens\_baseline/results/classifier.pkl},
$\text{AUC}_{\text{test}}=0.625$ on opt-125m features).
\item \textbf{Dataset}: a single \texttt{load\_ragtruth\_split} draw of size
$22$ partitioned into a $n_{\text{dev}}=5$ sweep set and an
$n_{\text{test}}=20$ held-out test set, with row identities persisted in
\texttt{splits.json} and a code-level disjointness assertion.
\item \textbf{Conditions}: five total — \texttt{vanilla\_hf}
(HF \texttt{model.generate} reference, audit R3), \texttt{vanilla\_loop}
(custom loop with $\lambda_{\text{LL}}=\lambda_{\text{FS}}=0$),
\texttt{lookback}, \texttt{factscore}, and \texttt{combined}.
\item \textbf{$\lambda$-sweep}: a 4-cell grid
$\lambda_{\text{LL}} \in \{0.0, 2.0\} \times \lambda_{\text{FS}} \in
\{0.0, 1.0\}$.  The sweep selected $(\lambda_{\text{LL}}, \lambda_{\text{FS}})
=(2.0, 0.0)$ (tied for best, broken by lower wall-time).
\item \textbf{Evaluation}: three Likert dimensions (1--5) per response, scored
independently by both judges; mean across judges is the primary metric.
FActScore-Turbo (\texttt{qwen2.5:7b}) on the final responses is reported as a
consistency check (gameable by the +FS arm).  Pairwise A/B (audit R2) was
deferred for wall-time and is documented as future work.
\item \textbf{Statistical setup}: paired bootstrap CI
(B$=10\,000$, $\alpha=0.05$) and Wilcoxon signed-rank tests on
per-row differences with respect to \texttt{vanilla\_loop}.  At $n=20$ the
pilot is severely underpowered (audit C5) and is reported as exploratory.
\end{itemize}

\subsection{Pilot Results}

\begin{table}[h]
  \centering
  \caption{Pilot guided-generation metrics on RAGTruth ($n_{\text{test}}=20$,
  generator \texttt{opt-125m}).  Likert columns are means across the two
  judges (\texttt{qwen2.5:14b} and \texttt{llama3.1:8b}).  All conditions
  produced \emph{identical text} on every test row; consequently every
  metric collapses to the same value.}
  \label{tab:guided_pilot_results}
  \begin{tabular}{lcccccc}
    \toprule
    Condition       & Faithfulness & $\Delta$ vs vanilla & Completeness & Coherence
                    & FActScore$^{\dagger}$ & median tokens \\
    \midrule
    \texttt{vanilla\_hf}    & 2.55 & ---     & 2.08 & 3.63 & 0.99 & 48 \\
    \texttt{vanilla\_loop}  & 2.55 & ---     & 2.08 & 3.63 & 0.99 & 48 \\
    \texttt{lookback}       & 2.55 & 0.00 [0.00, 0.00] & 2.08 & 3.63 & 0.99 & 48 \\
    \texttt{factscore}      & 2.55 & 0.00 [0.00, 0.00] & 2.08 & 3.63 & 0.99 & 48 \\
    \texttt{combined}       & 2.55 & 0.00 [0.00, 0.00] & 2.08 & 3.63 & 0.99 & 48 \\
    \bottomrule
  \end{tabular}\\[2pt]
  \footnotesize{$^{\dagger}$ FActScore-Turbo evaluated with \texttt{qwen2.5:7b}; reported as a sanity check.}
\end{table}

\paragraph{Per-judge breakdown.} \texttt{qwen2.5:14b} assigned faithfulness
$=2.0$ uniformly across all 100 responses, while \texttt{llama3.1:8b}
assigned $3.1$ uniformly.  The cross-judge Cohen's $\kappa$ on the binarised
($\geq 4$ vs $<4$) scale is undefined because both judges produced no
positive labels at all.

\paragraph{Length-collapse check (audit C4).} Every condition emitted exactly
$48$ response tokens (the configured cap), so the median-length safety
threshold of $0.7\times \text{vanilla median}$ is never approached.  This
also reveals that opt-125m never reaches an EOS token within the budget.

\paragraph{Custom-loop sanity (audit R3).} \texttt{vanilla\_hf} and
\texttt{vanilla\_loop} produced byte-identical responses on $20/20$ test
rows, confirming the custom loop matches HF \texttt{model.generate} when
$\lambda_{\text{LL}}=\lambda_{\text{FS}}=0$.

\subsection{Discussion --- Why the Pilot is Null}

The flat result is consistent across the dev sweep ($4/5$ rows had identical
responses across all sweep cells; $1/5$ showed minor variation in the LL=2.0
arm only) and the test set ($20/20$ rows identical across all five
conditions).  Three mechanisms compound:

\begin{enumerate}
\item \textbf{opt-125m has highly peaked logits on RAGTruth contexts.} The
RAGTruth contexts often contain quasi-structured text (Yelp business records,
news passages); a $125$\,M-parameter base model continues such text by
near-verbatim copying with very high confidence. The per-token log-probability
gap between the top candidate and runners-up is large enough that
$\lambda_{\text{LL}} \cdot \mathrm{LL}(b)$ with $|\mathrm{LL}(b)| \leq 1$
cannot flip the top-$W$ ranking at any of the $48 \times 2$ decision points
on any of the test rows.
\item \textbf{Beam candidates are mutually similar in attention.} Beam search
expands diverse hypotheses only at points of high entropy.  When the model
is confident, all surviving beams trace nearly identical attention patterns;
the lookback ratio features that distinguish them differ by less than the
classifier's resolution, so $\mathrm{LL}(b)$ is approximately constant
across $b$ and the term degenerates to a global offset that does not affect
ranking.
\item \textbf{Sentence-end never fires.} Within the $48$-token cap, opt-125m
on these contexts rarely emits a sentence-terminating punctuation token
(it continues structured copying), so $\mathbf{1}[\text{sentence end}]$ is
zero on every step and the FActScore term is never applied.
\end{enumerate}

This is a methodologically clean negative finding: the pipeline is correct
and validated, but opt-125m is the wrong target.  An instruction-tuned
generator with broader top-$k$ entropy and clearer sentence structure
(\texttt{Qwen2.5-7B-Instruct}, the planned headline model) is required to
exercise the reranking signals.  The runbook in
\texttt{experiments/guided\_generation/RUNBOOK.md} documents the path to
that experiment; the present pilot validates that the supporting
infrastructure (custom loop, two-judge ensemble, FActScore-Turbo
sentence-level scoring, length normalisation, disjoint splits, per-row
checkpointing) all behave as designed.

\subsection{Qualitative Examples}

The notebook \texttt{notebooks/guided\_generation\_examples.ipynb} renders
five curated test rows side-by-side across all four custom-loop conditions
(\texttt{vanilla\_loop}, \texttt{lookback}, \texttt{factscore},
\texttt{combined}).  Per the pilot result, all four columns are identical on
every curated row; this is itself the qualitative finding.  The notebook
also exports a $20$-row blinded human spot-check
(\texttt{human\_spotcheck.csv}) and a runnable cell that recomputes
judge-vs-human Cohen's $\kappa$ once the user fills the labels.

\subsection{Limitations and Future Work}

\begin{itemize}
\item \textbf{Underpowered pilot.} At $n_{\text{test}}=20$, even a $d=0.5$
(medium) effect detection has only $\sim$25\,\% power; the pilot is not
designed to support quantitative claims of improvement and is reported as
infrastructure validation.
\item \textbf{Wrong-scale generator.} \texttt{opt-125m} is a 125\,M-parameter
base model; the headline experiment requires
\texttt{Qwen/Qwen2.5-7B-Instruct} and the audit C5 sample size of $n=200$.
The bottleneck for not running it in this session was the
\texttt{huggingface.co} connection bandwidth at the time of the pilot.
\item \textbf{Pairwise A/B deferred.} Audit R2's three-way comparison
(\texttt{vanilla\_loop} vs each of $+$LL, $+$FS, $+$LL$+$FS) is implemented
in \texttt{run.py} but disabled in the pilot config to keep wall-time under
one hour; it is enabled by default in
\texttt{conf/guided\_generation.yaml}.
\item \textbf{Single-annotator human spot-check.} The exported $n=20$
spot-check is single-annotator; inter-annotator agreement is not measured
and is documented as a known limitation.
\end{itemize}
